---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
library(sf)
```

# lazysf

<!-- badges: start -->
<!-- badges: end -->

The goal of lazysf is to provide interactive delayed read of GDAL vector data sources. 

Vector data sources (a.k.a. "shapefiles") are files or web services or databases that provide tables of data fields. These fields
may include spatial geometry data (i.e. points, lines, polygons, and other planar types composed of coordinate paths). 

lazysf uses the dplyr/dbplyr 'tbl_lazy' mechanism by providing a GDAL-backend like many database packages in R.  The convenience function `lazysf()` provides a single-argument wrapper around the database-like workflows. 

See it in action!

```{r sf-action}
library(lazysf)
library(sf)
library(dplyr)

url <- "https://github.com/Nowosad/spData/raw/master/inst/shapes/NY8_bna_utm18.gpkg"
(x <- lazysf(url))
x %>% distinct(AREANAME) %>% arrange(AREANAME) 

plot(st_as_sf(x %>% 
                dplyr::filter(!(AREANAME %LIKE% "Ca%" | AREANAME %LIKE% "Bi%")) %>% 
                dplyr::select(AREANAME, geom)))
```

## Limitations

These are big promises!  It's very largely format dependent.  

We make no claims about performance or convenience, it will be affected by your system and your sf installation - lazysf just takes you closer the GDAL capabilities. 

For example, a query on a CSV, GeoJSON, or KML file local or remote is entirely
subject to the performance of the matching [GDAL driver](https://gdal.org/drivers/vector/index.html).

* big text files will be slow, they aren't a format suitable for database-like access
* geometry is not automatic (it *is* automatic for non-DB formats, with DB-formats it depends on your SQL)
* non-DB formats without a geometry column name will be called `_ogr_geometry_`, other non-DB formats like
ESRI's geodatabase have other names like `SHAPE`
* non-DB formats have access to special variable names and functions, listed on the [OGRSQL page](https://gdal.org/user/ogr_sql_dialect.html). Real DBs don't have these special OGRSQL features, but they do have their own. 

When using dplyr verbs (`filter()`, `select()`, `mutate()`, `transmute()`,
`arrange()`, `left_join()`, ...) we are also subject to the rules of SQL
translation. There are no specific ones provided by lazysf but that might
change.

Wrappers around lazysf could provide more specific tools for particular formats. 

### Can't we just do this with sf's query argument?

Yes (actually that is what lazysf uses) but with sf alone you get a fully
materialized sf data frame, so you better get that query right first time!

With lazysf you get some control over intermediate steps, potentially expensive
queries will only be run for a preview of the data until you are ready to fetch
it.

## Installation

You can install the dev version of lazysf from [GitHub](https://github.com/mdsumner/lazysf) with:

``` r
remotes::install_github("mdsumner/lazysf")
```

## Example

This is a basic example. 

```{R basic}
library(lazysf)
f <- system.file("gpkg/nc.gpkg", package = "sf", mustWork = TRUE)

## specify only the data source
lazysf(f)

## specify the data source and a query to run
lazysf(f, query = "SELECT AREA, FIPS, geom FROM \"nc.gpkg\" WHERE AREA < 0.1")

## specify the data source and the table/layer to access
lazysf(f, layer = "nc.gpkg") %>% 
  dplyr::select(AREA, FIPS, geom) %>% 
  dplyr::filter(AREA < 0.1)


## above was a real database (Geopackage), now with an actual shapefile
shp <- lazysf(system.file("shape/nc.shp", package = "sf", mustWork = TRUE))
library(dplyr)
shp %>%
 filter(NAME %LIKE% 'A%') %>%
 mutate(abc = 1.3) %>%
 select(abc, NAME, `_ogr_geometry_`) %>%
 arrange(desc(NAME))  #%>% show_query()
```

Online sources can also work if your build of sf supports. 

```{r online}
# online sources can work
geojson <- file.path("https://raw.githubusercontent.com/SymbolixAU",
                      "geojsonsf/master/inst/examples/geo_melbourne.geojson")
lazysf(geojson)
```


Also works on PostgreSQL with spatial PostGIS (or just normal postgres).  

Note that the connection object is a transparent string so take care with your
credentials and don't share in public (currently trying to work out if only the
externalptr should be kept in this, or in all cases with GDAL sources).


The `SFSQL_PG()` driver is a convenience wrapper around the `SFSQL()` driver. 

```{r postgres}
con <- dbConnect(SFSQL_PG(),
          host = Sys.getenv("dbedu_host"),
          dbname = Sys.getenv("dbedu_dbname"), 
          user = Sys.getenv("dbedu_usr"), 
          password = Sys.getenv("dbedu_pwd"))

(sfx <- lazysf(con, query = "SELECT st_area(st_transform(geom, '+proj=laea +lon_0=-68 +lat_0=-25')) / 1e6 AS sq_km, 
                                        codigo_comuna, geom FROM mapa_comunas WHERE codigo_region = '03'"))
```

You can construct the database connection string yourself and provide it
directly to lazysf() if you want. See [GDAL Postgres](https://gdal.org/drivers/vector/pg.html) 
for details on this format, and this capability is available for any of the GDAL formats (see limitations above).  



---
## Code of Conduct

Please note that the lazysf project is released with a [Contributor Code of
Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By
contributing to this project, you agree to abide by its terms.
